Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-01-26 SCP2008/paper/metamonads.tex
cleaned up the biblography, refreshed a few references.

Therefore, in MetaOCaml, code must be generated just right (see
\cite{TahaThesis} for many simple examples).  For more complex
examples, new techniques are necessary, e.g., abstract interpretation
\cite{KiselyovTaha}.  But more problems remain
\cite{Padua:MetaOcaml:04}: generating binding forms (``names'')
when generating loop bodies or conditional branches, and making
continuation-passing style (CPS) code clear.  Many authors
understandably shy away from CPS code as it quickly becomes
unreadable.  But this is needed for proper name generation.
The problems of compositionality of code generators, expressing
dependencies among them and domain-specific knowledge remain.

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
\footnote{We describe here a new version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have also changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of LU decomposition 
and linear system solving as our running examples to demonstrate our
techniques.  Specifically, our contributions are:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{KiselyovTaha,MSP:PADL04} for generating control structures
      such as loops and conditionals. The extension is non-trivial
	  because of control dependencies and because let-insertion, as we argue,
	  is a control effect on its own: for example
      \texttt{let x = exp in \dots} has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

@@

The solution to this problem is to use continuation-passing style (CPS). Its
benefits were first pointed out by \cite{Bondorf:92} in the context of partial
evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
code generation. Like \cite{conf/pepm/BondorfD94}, we use this in the 
context of writing a cogen by hand.  Now, |param_code2 plus one| gives us the
desired code.

@@

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
%
We first instantiate the container |GVC_I| for a vector over an
integer domain. We pass the container to the main functor |GenLA|,
since all linear algebra codes are parameterized by the
container and the domain included in it. We open the |G_GVC_I| module
to make all its components easily available. One of those components
is the GE generator, |GE| (which we open as well). Besides |GE|, the
module |G_GVC_I| includes the generator for GE solvers.  The main
generator functor is |GenGE|, shown later.  We instantiate the
generator by passing to the functor the ``record'' of various aspects; the
order is irrelevant, but all aspects such as |Det| must be
specified. In the code above, we request GE generation with full
pivot, fraction-free update, operate on non-augmented matrix and
return the |U| factor, determinant and the rank. We run the monad
passing the initial state |[]| and obtain the code, which we can see
by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
or with off-shoring. The code for |resIV5|
(App.~\ref{app:code-GenIV5}) shows full
pivoting, determinant and rank tracking. The code for all these
aspects is fully inlined; no extra functions are invoked and no tests
other than those needed by the LU algorithm itself are performed. The
resulting function returns a triple |int array * int * int| of the
U-factor, determinant and the rank.

@@

The monad in this paper is similar to the one described in
\cite{MSP:PADL04,KiselyovTaha}.  However those papers used only
|retN| and fixpoints (for generation-time iterations).  This paper
does not involve monadic fixpoints because the generator is not
recursive, but heavily relies on monadic operations for generating
conditionals and loops.

@@

There are many more aspects which can also be handled:
error reporting (i.e. asking for the determinant of a 
non-square matrix), memory hierarchy issues, loop-unrolling
\cite{Padua:MetaOcaml:04},
warnings when zero-testing is undecidable and
a value is only probabilistically non-zero, etc.

@@

% Embed .bbl file directly to make document self-contained
% \bibliography{metamonads}
% \bibliographystyle{elsart-num}

\begin{thebibliography}{10}
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi

\bibitem{Carette06}
J.~Carette, Gaussian {E}limination: a case study in efficient genericity with
  {MetaOCaml}, Science of Computer Programming 62~(1) (2006) 3--24, special
  Issue on the First MetaOCaml Workshop 2004.

\bibitem{Monagan:2001:M7P}
M.~B. Monagan, K.~O. Geddes, K.~M. Heal, G.~Labahn, S.~M. Vorkoetter,
  J.~McCarron, P.~DeMarco, Maple 7 Programming Guide, Waterloo Maple Inc.,
  2001.

\bibitem{Gruntz:1994:IG}
D.~Gruntz, M.~Monagan, Introduction to {Gauss}, \textsc{SIGSAM Bulletin}:
  Communications on Computer Algebra 28~(2) (1994) 3--19.

\bibitem{Axiom}
R.~D. Jenks, R.~S. Sutor, {AXIOM}: The Scientific Computation System, Springer
  Verlag, 1992.

\bibitem{Czarnecki}
K.~Czarnecki, U.~W. Eisenecker, Generative programming: methods, tools, and
  applications, ACM Press/Addison-Wesley Publishing Co., 2000.

\bibitem{Veldhuizen:1998:ISCOPE}
T.~L. Veldhuizen, Arrays in {Blitz++}, in: Proceedings of the 2nd International
  Scientific Computing in Object-Oriented Parallel Environments (ISCOPE'98),
  Lecture Notes in Computer Science, Springer-Verlag, 1998.

\bibitem{musser89generic}
D.~R. Musser, A.~A. Stepanov, Generic programming, in: {ISSAC 1988}:
  Proceedings of the International Symposium on Symbolic and Algebraic
  Computation, Vol. 358 of Lecture Notes in Computer Science, Springer-Verlag,
  1989, pp. 13--25.

\bibitem{musser94algorithmoriented}
D.~R. Musser, A.~A. Stepanov, Algorithm-oriented generic libraries, Software -
  Practice and Experience 24~(7) (1994) 623--642.

\bibitem{BOOST}
J.~Siek, L.-Q. Lee, A.~Lumsdaine, The Boost Graph Library: User Guide and
  Reference Manual, Addison-Wesley, 2002.

\bibitem{POOMA}
I.~John V.W.~Reynders, J.~C. Cummings, The {POOMA} framework, Comput. Phys.
  12~(5) (1998) 453--459.

\bibitem{ATLAS}
R.~C. Whaley, A.~Petitet, J.~J. Dongarra, Automated empirical optimization of
  software and the {ATLAS} project, Parallel Computing 27~(1--2) (2001) 3--35.

\bibitem{HygienicMacros}
E.~E. Kohlbecker, D.~P. Friedman, M.~Felleisen, B.~F. Duba, Hygienic macro
  expansion., in: LISP and Functional Programming, 1986, pp. 151--161.

\bibitem{camlp4}
D.~de~Rauglaudre, Camlp4 reference manual,
  \url{http://caml.inria.fr/camlp4/manual/} (Jan. 2002).

\bibitem{conf/dagstuhl/CzarneckiOST03}
K.~Czarnecki, J.~T. O'Donnell, J.~Striegnitz, W.~Taha, {DSL} implementation in
  {MetaOCaml}, {Template Haskell}, and {C++}., in: C.~Lengauer, D.~S. Batory,
  C.~Consel, M.~Odersky (Eds.), Domain-Specific Program Generation, Vol. 3016
  of Lecture Notes in Computer Science, Springer, 2003, pp. 51--72.

\bibitem{CTHL03}
C.~Calcagno, W.~Taha, L.~Huang, X.~Leroy, Implementing multi-stage languages
  using asts, gensym, and reflection, in: K.~Czarnecki, F.~Pfenning,
  Y.~Smaragdakis (Eds.), Generative Programming and Component Engineering
  (GPCE), LNCS, Springer-Verlag, 2003.

\bibitem{metaocaml-org}
{MetaOCaml}, \url{http://www.metaocaml.org}.

\bibitem{TahaSheard97}
W.~Taha, T.~Sheard, Multi-stage programming with explicit annotations, in:
  Proceedings of the Symposium on Partial Evaluation and Semantic-Based Program
  Manipulation {(PEPM)}, ACM Press, Amsterdam, 1997, pp. 203--217.

\bibitem{TahaThesis}
W.~Taha, Multi-stage programming: its theory and applications, Ph.D. thesis,
  Oregon Graduate Institute of Science and Technology (1999).

\bibitem{Taha2000}
W.~Taha, A sound reduction semantics for untyped {CBN} multi-stage computation.
  {Or}, the theory of {MetaML} is non-trival., in: PEPM, 2000, pp. 34--43.

\bibitem{Pueschel:05}
M.~P{\"u}schel, J.~M.~F. Moura, J.~Johnson, D.~Padua, M.~Veloso, B.~W. Singer,
  J.~Xiong, F.~Franchetti, A.~Ga\v{c}i\'{c}, Y.~Voronenko, K.~Chen, R.~W.
  Johnson, N.~Rizzolo, {SPIRAL}: Code generation for {DSP} transforms,
  Proceedings of the IEEE, special issue on "Program Generation, Optimization,
  and Adaptation" 93~(2).

\bibitem{Kennedy01Telescoping}
K.~Kennedy, B.~Broom, K.~Cooper, J.~Dongarra, R.~Fowler, D.~Gannon,
  L.~Johnsson, J.~Mellor-Crummey, L.~Torczon, Telescoping languages: A strategy
  for automatic generation of scientific problem-solving systems from annotated
  libraries, Journal of Parallel and Distributed Computing 61~(12) (2001)
  1803--1826.

\bibitem{dongarra7}
Z.~Chen, J.~Dongarra, P.~Luszczek, K.~Rothe, Lapack for clusters project: An
  example of self adapting numerical software, Hawaii International Conference
  on System Sciences HICSS-37.

\bibitem{Veldhuizen:2004}
T.~L. Veldhuizen, Active libraries and universal languages, Ph.D. thesis,
  Indiana University Computer Science (May 2004).
\newline\urlprefix\url{\url{http://osl.iu.edu/~tveldhui/papers/2004/dissertati%
on.pdf}}

\bibitem{KiselyovTaha}
O.~Kiselyov, K.~N. Swadi, W.~Taha, A methodology for generating verified
  combinatorial circuits, in: EMSOFT '04: Proceedings of the fourth ACM
  international conference on Embedded software, ACM Press, New York, NY, USA,
  2004, pp. 249--258.

\bibitem{Padua:MetaOcaml:04}
A.~Cohen, S.~Donadio, M.~J. Garzara'n, C.~Herrmann, D.~Padua, In search for a
  program generator to implement generic transformations for high-performance
  computing, metaOCaml Workshop, October 2004.

\bibitem{CaretteKiselyov05}
J.~Carette, O.~Kiselyov, Multi-stage programming with {F}unctors and {M}onads:
  eliminating abstraction overhead from generic code, in: Generative
  Programming and Component-based Engineering {GPCE}, 2005, pp. 256--274.

\bibitem{MSP:PADL04}
J.~L. Eckhardt, R.~Kaiabachev, K.~N. Swadi, W.~Taha, O.~Kiselyov, Practical
  aspects of multi-stage programming, rice University Techical Report
  {TR05}-451,
  \url{http://www.cs.rice.edu/~taha/publications/preprints/2004-02-16.pdf}.

\bibitem{metamonadsURL}
{Source code}, \url{http://www.cas.mcmaster.ca/~carette/metamonads/}.

\bibitem{Golub-vanLoan}
G.~H. {Golub}, C.~F. {Van Loan}, Matrix Computations, 3rd Edition, Johns
  Hopkins University Press, Baltimore, MD, 1996.

\bibitem{kiczales97aspectoriented}
G.~Kiczales, J.~Lamping, A.~Menhdhekar, C.~Maeda, C.~Lopes, J.-M. Loingtier,
  J.~Irwin, Aspect-oriented programming, in: M.~Ak\c{s}it, S.~Matsuoka (Eds.),
  Proceedings European Conference on Object-Oriented Programming, Vol. 1241,
  Springer-Verlag, Berlin, Heidelberg, and New York, 1997, pp. 220--242.

\bibitem{709568}
J.~Irwin, J.-M. Loingtier, J.~R. Gilbert, G.~Kiczales, J.~Lamping,
  A.~Mendhekar, T.~Shpeisman, Aspect-oriented programming of sparse matrix
  code, in: ISCOPE '97: Proceedings of the Scientific Computing in
  Object-Oriented Parallel Environments, Springer-Verlag, London, UK, 1997, pp.
  249--256.

\bibitem{mendhekar97rg}
A.~Mendhekar, G.~Kiczales, J.~Lamping, {RG}: {A} case-study for aspect-oriented
  programming, Tech. Rep. SPL97-009 P9710044, Palo Alto, CA, USA (February
  1997).

\bibitem{journals/cacm/parnas72a}
D.~L. Parnas, On the criteria to be used in decomposing systems into modules.,
  Commun. ACM 15~(12) (1972) 1053--1058.

\bibitem{journals/tse/Parnas76}
D.~L. Parnas, On the design and development of program families., IEEE Trans.
  Software Eng. 2~(1) (1976) 1--9.

\bibitem{EWD:EWD447}
E.~W. Dijkstra, On the role of scientific thought, published as
  \cite{EWD:EWD447pub}.
\newline\urlprefix\url{\url{http://www.cs.utexas.edu/users/EWD/ewd04xx/EWD447.%
PDF}}

\bibitem{ZhCaJeMo06a}
W.~Zhou, J.~Carette, D.~J. Jeffrey, M.~B. Monagan, Hierarchical representations
  with signatures for large expression management, in: Proceedings of
  Artificial Intelligence and Symbolic Computation, Vol. 4120 of Lecture Notes
  in Computer Science, 2006, pp. 254--268.

\bibitem{Bondorf:92}
A.~Bondorf, Improving binding times without explicit {CPS}-conversion, in: 1992
  ACM Conference on Lisp and Functional Programming. San Francisco, California,
  1992, pp. 1--10.

\bibitem{conf/pepm/BondorfD94}
A.~Bondorf, D.~Dussart, Improving {CPS}-based partial evaluation: Writing cogen
  by hand., in: PEPM, 1994, pp. 1--9.

\bibitem{Haskell98Report}
S.~Peyton~Jones, et~al., The Revised Haskell 98 Report, Cambridge Univ.\null{}
  Press, 2003, also on \url{http://haskell.org/}.

\bibitem{moggi-notions}
E.~Moggi, Notions of computation and monads, Information and Computation 93~(1)
  (1991) 55--92.

\bibitem{Filinski:Representing}
A.~Filinski, Representing monads., in: POPL, 1994, pp. 446--457.

\bibitem{liang-interpreter}
S.~Liang, P.~Hudak, M.~Jones, Monad transformers and modular interpreters, in:
  {POPL} '95: Conference Record of the Annual {ACM} Symposium on Principles of
  Programming Languages, {ACM} {P}ress, {N}ew {Y}ork, 1995, pp. 333--343.

\bibitem{taha-environment}
W.~Taha, M.~F. Nielsen, Environment classifiers, in: {POPL} '03: Conference
  Record of the Annual {ACM} Symposium on Principles of Programming Languages,
  {ACM} {P}ress, {N}ew {Y}ork, 2003, pp. 26--37.

\bibitem{mlton-proplist}
{{MLton} team}, Property list, \url{http://mlton.org/PropertyList}.

\bibitem{Watt:2002:HCA}
S.~M. Watt, Aldor, in: J.~Grabmeier, E.~Kaltofen, V.~Weispfennig (Eds.),
  Computer Algebra Handbook: Foundations, Applications, Systems, Springer
  Verlag, 2003.

\bibitem{Gluck95}
R.~Gl{\"u}ck, R.~Nakashige, R.~Z{\"o}chling, Binding-time analysis applied to
  mathematical algorithms, in: System Modelling and Optimization, 1995.

\bibitem{GluckJ97}
R.~Gl{\"u}ck, J.~J{\o}rgensen, An automatic program generator for multi-level
  specialization., Lisp and Symbolic Computation 10~(2) (1997) 113--158.

\bibitem{edsl}
P.~Hudak, Building domain-specific embedded languages, ACM Comput. Surv.
  28~(4es) (1996) 196.

\bibitem{WhenHowDSL}
M.~Mernik, J.~Heering, A.~M. Sloane, When and how to develop domain-specific
  languages, ACM Comput. Surv. 37~(4) (2005) 316--344.

\bibitem{EWD:EWD447pub}
E.~W. Dijkstra, On the role of scientific thought, in: Selected Writings on
  Computing: A Personal Perspective, Springer-Verlag, 1982, pp. 60--66.

\end{thebibliography}
\appendix
\section{The \texttt{perform} monad notation}
\label{app:perform}
We support four different constructs to introduce a monadic
expressions:
\begin{code}
  perform exp
  perform exp1; exp2
  perform x <-- exp1; exp2
  perform let x = foo in exp
\end{code}
which is almost literally the grammar of the Haskell's "do"-notation,
with the differences that Haskell uses |do| and |<-| where we use
|perform| and |<--|.
We support not only |let x = foo in ...|  expressions but arbitrarily
complex let-expressions, including |let rec| and |let module|.

>>>>

Therefore, in MetaOCaml, code must be generated just right (see
\cite{TahaThesis} for many simple examples).  For more complex
examples, new techniques are necessary, e.g., abstract interpretation
\cite{KiselyovTaha}.  But more problems remain
\cite{scp-CohenDGHKP06}: generating binding forms (``names'')
when generating loop bodies or conditional branches, and making
continuation-passing style (CPS) code clear.  Many authors
understandably shy away from CPS code as it quickly becomes
unreadable.  But this is needed for proper name generation.
The problems of compositionality of code generators, expressing
dependencies among them and domain-specific knowledge remain.

In this paper, we report on our continued progress \cite{CaretteKiselyov05}
\footnote{We describe here a new version of our generator
  dealing with the complete LU decomposition algorithm, as well as
  linear solving. We worked out previously missing aspects of in-place
  updates, representing permutation matrices, dealing with augmented 
  input matrix, and back-propagation. We have also changed the representation
  of domain-specific knowledge about permissible compositions of aspects.
  Also included is a careful description of all the aspects involved, as
  well as documenting our development methodology for highly parametric
  scientific software.}
in using code generation for scientific (both numeric and symbolic)
software.  We will use the algorithm family of LU decomposition 
and linear system solving as our running examples to demonstrate our
techniques.  Specifically, our contributions are:
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{KiselyovTaha,SwadiMonadic06} for generating control structures
      such as loops and conditionals. The extension is non-trivial
	  because of control dependencies and because let-insertion, as we argue,
	  is a control effect on its own: for example
      \texttt{let x = exp in \dots} has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      the |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{insure composability of aspects} even
      for aspects that use state and have to be accounted for in many
      places in the generated code.
    \item Encode domain-specific knowledge in the generators which 
      will catch domain-specific instantiation errors at generation
      time.
    \item Provide a thorough breakdown of the family of LU decomposition
      algorithms.
\end{itemize}

@@

The solution to this problem is to use continuation-passing style (CPS). Its
benefits were first pointed out by \cite{Bondorf:92} in the context of partial
evaluation, and extensively used by \cite{SwadiMonadic06,KiselyovTaha} for
code generation. Like \cite{conf/pepm/BondorfD94}, we use this in the 
context of writing a cogen by hand.  Now, |param_code2 plus one| gives us the
desired code.

@@

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
\end{code}
%
We first instantiate the container |GVC_I| for a vector over an
integer domain. We pass the container to the main functor |GenLA|,
since all linear algebra codes are parameterized by the
container and the domain included in it. We open the |G_GVC_I| module
to make all its components easily available. One of those components
is the GE generator, |GE| (which we open as well). Besides |GE|, the
module |G_GVC_I| includes the generator for GE solvers.  The main
generator functor is |GenGE|, shown later.  We instantiate the
generator by passing to the functor the ``record'' of various aspects; the
order is irrelevant, but all aspects such as |Det| must be
specified. In the code above, we request GE generation with full
pivot, fraction-free update, operate on non-augmented matrix and
return the |U| factor, determinant and the rank. We run the monad
passing the initial state |[]| and obtain the code, which we can see
by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
or with offshoring\cite{offshoring}. The code for |resIV5|
(App.~\ref{app:code-GenIV5}) shows full
pivoting, determinant and rank tracking. The code for all these
aspects is fully inlined; no extra functions are invoked and no tests
other than those needed by the LU algorithm itself are performed. The
resulting function returns a triple |int array * int * int| of the
U-factor, determinant and the rank.

@@

The monad in this paper is similar to the one described in
\cite{SwadiMonadic06,KiselyovTaha}.  However those papers used only
|retN| and fixpoints (for generation-time iterations).  This paper
does not involve monadic fixpoints because the generator is not
recursive, but heavily relies on monadic operations for generating
conditionals and loops.

@@

There are many more aspects which can also be handled:
error reporting (i.e. asking for the determinant of a 
non-square matrix), memory hierarchy issues, loop-unrolling
\cite{scp-CohenDGHKP06},
warnings when zero-testing is undecidable and
a value is only probabilistically non-zero, etc.

@@

% Embed .bbl file directly to make document self-contained
\bibliography{metamonads}
\bibliographystyle{elsart-num}

%% \begin{thebibliography}{10}
%% \expandafter\ifx\csname url\endcsname\relax
%%   \def\url#1{\texttt{#1}}\fi
%% \expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi

\appendix
\section{The \texttt{perform} monad notation}
\label{app:perform}
We support four different constructs to introduce a monadic
expressions:
\begin{code}
  perform exp
  perform exp1; exp2
  perform x <-- exp1; exp2
  perform let x = foo in exp
\end{code}
which is almost literally the grammar of the Haskell's "do"-notation,
with the differences that Haskell uses |do| and |<-| where we use
|perform| and |<--|.
We support not only |let x = foo in ...|  expressions but arbitrarily
complex let-expressions, including |let rec| and |let module|.
